---
bibliography: Whitebib.bib
csl: ecology-letters.csl
editor_options:
  chunk_output_type: console
fontsize: 12pt
geometry: margin=1in
header-includes: \usepackage{float} \usepackage{lineno} \usepackage{setspace}\doublespacing
  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount}
  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite}
  \usepackage{fancyhdr} \pagestyle{fancy} \fancyhf{} \fancyfoot[LE,LO]{\textcolor{red}{Preprint
  - This work has not yet been peer-reviewed}} \fancyfoot[RE,RO]{\thepage} \renewcommand{\headrulewidth}{0pt}
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: no
  word_document: default
  html_document:
    df_print: paged
---




\begin{center}
\textbf{\Large State-level variation for initial COVID-19 dynamics in the United States: The role of local government interventions}
\vspace{7 mm}
	
\textsc{Easton R. White\footnote{Easton.White@uvm.edu}$^{,2}$, Laurent H\'{e}bert-Dufresne $^{3,4}$}
\vspace{5 mm}

\normalsize{\indent $^2$Department of Biology, University of Vermont, VT 05405, USA \\ {$^3$} Vermont Complex Systems Center, University of Vermont, VT 05405, USA \\ $^4$Department of Computer Science, University of Vermont, VT 05405, USA}
\end{center}


\vspace{5 mm}

\textbf{Abstract}

In the development of an epidemic, metrics such as $R_0$, doubling time, and case fatality rates are important in understanding and predicting the course of an epidemic. However, if collected over country or regional scales, these metrics hide important smaller-scale, local dynamics. We examine how commonly used epidemiological metrics differ for each individual state within the United States during the initial COVID-19 outbreak. We found that the case number, and trajectory of cases, differs considerably between states. We show that early non-pharmaceutical, government actions, were the most important determinant of epidemic dynamics. Although individual states are clearly not independent, they can serve as mini, natural experiments in how different demographic patterns and government responses can impact the course of an epidemic. Thus, these results should be used to better understand, in near real-time, what actions are working most effectively. 


\vspace{3 mm}

Keywords: SARS-CoV-2, COVID-19, spatial heterogeneity, doubling time

\vspace{3 mm}

Daily updates to figures in this manuscript are available at: https://github.com/eastonwhite/COVID19_US_States


```{r load_packages,echo=F,warning=F,message=F}
if (!require("pacman",character.only = TRUE))
  {
    install.packages("pacman",dep=TRUE)
    if(!require("pacman",character.only = TRUE)) stop("Package not found")
  }

# Keeping below source for github package. Ask Easton whether pacman works for github packages or not.
#devtools::install_github("rensa/stickylabeller")
pacman::p_load(patchwork, dplyr,tidyr,ggplot2,ggrepel,viridis,usmap)
```


```{r, echo=F,message=F,warning=F}
# Load data for numbers in introduction
covid <- read.csv('https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv',header=T)



require(scales)
#sum(covid[,ncol(covid)])
#length(table(covid$Country.Region))
#format(as.Date(substr(names(covid)[ncol(covid)],2,8),format = '%m.%d.%y'),'%B %d')
```

\clearpage

# Introduction

The global SARS-CoV-2 (COVID-19) pandemic began in Wuhan, China in late 2019 [@WHO2020]. As of `r paste(format(as.Date(substr(names(covid)[ncol(covid)-1],2,8),format = '%m.%d.%y'),'%B %d'),'th',sep='')`, `r comma(sum(covid[,ncol(covid)-1]))` cases have been reported across `r length(table(covid$Country.Region))` countries and regions. There has been several sets of efforts to track the progression of the outbreak across the world and within countries. For example, John Hopkins University Center for Systems Science and Engineering (CSSE) has compiled data from various sources, including the US Center for Disease Control and the World Health Organization, to present a global picture of COVID-19 cases and deaths [@Dong2020]. These efforts have allowed for international scientific research and political decision-making. Although data are collected at local scales (e.g. within hospitals), in an emerging pandemic data is typically reported at the country level. This allows for interesting comparisons between countries [@Anderson2020,@Jombart2020] and for information from an earlier affected country to be used to slow the outbreak in other places. For instance, South Korea was able to flatten their outbreak curve through early and widespread testing as well as strict quarantine policies (citation). However, country-level analyses still hide more local dynamics that are important to the overall epidemic progression.

Spatial heterogeneity is important for population dynamics generally [@Levin1992,@Hanksi2001,@Schreiber2010] and in particular for understanding the progression of infectious disease dynamics [@Grenfell1995,@Park2012]. Spatial heterogeneity can include differences in local population density, movement patterns, suitability of environmental conditions for transmission, among other factors. For instance, @Keeling2001 showed how spatial distribution and size of farms affected the 2001 UK Foot and Mouth Epidemic.

<!--
For the current COVID-19 outbreak, researchers have noted the different trajectories even within countries. For example, in China... (citations). MORE TEXT HERE.
-->

Here we examine the progression of COVID-19 for state-level differences within the United States. We examine how commonly-used metrics, including doubling time, can vary state to state and compared to the US as a whole. Although not independent units, we can use state-level data to understand the progression of the outbreak across different replicates within a country. We then show that demography, education, wealth, and other hypotheses were poor predictors of doubling time. Instead, we show that doubling time was more tightly correlated with state-level governmental actions, including restricting businesses. We also provide a link to daily updates of our findings as the exact quantitative results are likely to change over the course of the epidemic. 

# Results and Discussion



<!--Load data and start analyses-->

```{r updated_data_cleaning,echo=F,eval=T,warning=F,message=F}
# Load data here automatically
require(dplyr)
require(tidyr)

source('CompileTimeSeries.R')

`%notin%` <- Negate(`%in%`)
# Subset data just for the USA
covid <- covid %>%
  filter(Country.Region == 'US') %>%
  filter(!grepl(',', Province.State)) %>% 
  filter(Province.State %notin% c('Unassigned Location (From Diamond Princess)','Grand Princess Cruise Ship','Wuhan Evacuee','US','Diamond Princess','Recovered','Grand Princess'))

# Change data to long format and adjust date variable
covid <- covid %>%
  #gather('Date','Cases',-Province.State,-Country.Region,-Lat,-Long) %>%
  #mutate(Date = sub('.',replacement = '',Date)) %>%
  mutate(Date = as.Date(Date,format='%m-%d-%y')) %>%
  mutate(DayOfYear = as.numeric(strftime(Date, format = "%j")))
```


```{r cases_vs_time,echo=F,warning=F,message=F, fig.height=4,eval=T,fig.cap='(Left panel) Cases versus time for the whole United States. (Right panel) Log number of cases versus time for the whole United States. The red, dashed line is the line of best fit for all the data and the blue, solid line is the line of best fit since Feburary 29th. \\label{fig:cases_vs_time}'}
# Basic plots of cases over time and broken up by state (this might be in the methods section)

source('rollingmean.R')
US_cases_by_day <- covid %>%
  group_by(Date) %>%
  summarize(total_cases = sum(Cases)) %>%
  mutate(DayOfYear=as.numeric(strftime(Date, format = "%j"))-min(as.numeric(strftime(Date, format = "%j"))))



#jpeg('cases_vs_time-1.jpeg',width = 7, height = 5, units='in',res=500)
par(mfrow=c(1,3),oma=c(1,0.5,0.5,0.5),mar=c(4,4,0.5,0.5))
plot(US_cases_by_day$Date,US_cases_by_day$total_cases,las=1,type='o',pch=16, ylab = 'Total number of US cases',xlab='')
#points(US_cases_by_day$Date,0.1875+0.11667*US_cases_by_day$Date)
#US_cases_by_day = US_cases_by_day[US_cases_by_day$total_cases>0,]
#dude=summary(lm(log(US_cases_by_day$total_cases)~US_cases_by_day$Date))
plot(US_cases_by_day$Date,US_cases_by_day$total_cases,las=1,type='p',pch=16, ylab = ' ',xlab='',log='y')
mtext(text = 'Date',side = 1,line = -1,outer = T,cex = 1.4,adj = 0.55)
fit_lm <- lm(log(US_cases_by_day$total_cases)~US_cases_by_day$Date)
points(US_cases_by_day$Date,exp(fit_lm$fitted.values),type='l',lwd=2,col=2,lty=2)

require(tidyquant)

rollmean <- US_cases_by_day %>%
  tq_mutate(
    # tq_mutate args
    select     = total_cases,
    mutate_fun = rollapply, 
    width      = 7,
    align      = "right",
    FUN        = calc_slope,
    col_rename = "mean_7"
  ) 


plot(rollmean$Date,log(2)/rollmean$mean_7,type='o',pch=16,las=1,ylab='Doubling time (7-day windows)',xlab='Date',cex.lab=1.2,ylim=c(1.8,6))


# Plot regression line for data starting on Feb 29th
#fit_lm2 <- lm(log(US_cases_by_day$total_cases[39:nrow(US_cases_by_day)])~US_cases_by_day$Date[39:nrow(US_cases_by_day)])
#points(US_cases_by_day$Date[39:nrow(US_cases_by_day)],exp(fit_lm2$fitted.values),type='l',lwd=2,lty=1,col='blue')
#dev.off()

```

We used data compiled by John Hopkins University Center for Systems Science and Engineering [@Dong2020]. The United States has seen exponential growth in the number of cases, especially since February 29th (Fig. \ref{fig:cases_vs_time}). Country-level results, however, hide underlying dynamics within each state. 

Therefore, we examined how the number of cases changed over time within each state. To properly compare the progression of the epidemic across states, we looked at the log number of cases since the first day a state reported 25 cases (Fig. \ref{fig:measurements_by_states}). On a log scale, a straight line of the cases over time indicates exponential growth where the slope of the line is the exponential growth parameter. 

## State-level variation in COVID-19 trajectories

We found considerable differences between states in how the outbreak has progressed (Fig. \ref{fig:measurements_by_states}). States like New York, New Jersey, and Michigan have experienced a doubling of cases approximately every two days. Conversely, Massachusetts has experienced a doubling time closer to five days. We mapped doubling time across the US and found regional differences where the West Coast and Northeast have seen large doubling times, i.e. slower outbreak dynamics (Fig. \ref{fig:measurements_by_states_map}).


```{r measurements_by_states,echo=F,warning=F,message=F, fig.height=5,eval=T,fig.cap='The log number of cases over time for each individual state that recorded more than 25 cases over at least three days. The light grey diagonal lines represent the growth trajectory for doubling times of 2, 4, and 10 days. The log number of the starting value (intial number of cases on first day when at least 25 cases were recorded) had to be subtracted on the y-axis to standarize the graph across states. \\label{fig:measurements_by_states}'}

# Filter out local county data and focus on states with at least 3 data points of 25 cases or more
covid_by_state <- covid %>%
  group_by(Date,Province.State) %>%
  summarise(Cases = sum(Cases),Deaths=sum(Deaths),Recovered = sum(Recovered), DayOfYear = mean(DayOfYear)) %>%
  filter(Cases > 24) %>%
  group_by(Province.State) %>%
  filter(length(Cases)>3)

# Record first date of more than 25 cases
time_of_25plus_cases <- covid_by_state %>%
  group_by(Province.State) %>%
  filter(length(Cases)>3) %>%
  summarize(first_confirmed = Date[which(Cases>24)[1]])

covid_by_state = left_join(covid_by_state,time_of_25plus_cases,by="Province.State")


starting_number = covid_by_state %>%
  group_by(Province.State) %>%
  summarize(starting_value = min(Cases))
covid_by_state <- left_join(covid_by_state,starting_number)

covid_by_state$DaysSince25 <- as.numeric(as.Date(covid_by_state$Date,format='%m.%d.%y')-as.Date(covid_by_state$first_confirmed,format='%m.%d.%y'))

require(ggplot2) 
require(ggrepel)
require(viridis)

states <- ggplot(data=covid_by_state,aes(x=DaysSince25, y=(log(Cases)) - (log(starting_value)),color=Province.State))  + 
  geom_abline(slope = log(2)/2, intercept = 0,color='lightgrey',size=2) +
  geom_abline(slope = log(2)/4, intercept = 0,color='lightgrey',size=2) +
  geom_abline(slope = log(2)/10, intercept = 0,color='lightgrey',size=2) +
  scale_color_viridis(discrete=T,begin = 0.1,end=0.9,option='viridis') +
  geom_line(alpha=0.5) + 
  theme_bw() + theme(legend.position = "none",axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14)) + 
  geom_label_repel(data = subset(covid_by_state, DayOfYear== max(DayOfYear)),aes(label = Province.State),nudge_x = 0.1,na.rm = TRUE,segment.color = NA,fill=NA,label.size = NA,box.padding = 0) + 
  coord_cartesian(xlim =c(0.5,24),ylim =c(-0.1,7)) +
  ylab('log (Cases) - log (Starting value)') + xlab('Days since 25 cases recorded') + 
  annotate("text", x=14, y=5.2, label='doubling every 2 days', size=3) +
  annotate("text", x=18, y=3, label='doubling every 4 days', size=3) + 
  annotate("text", x=15, y=1, label='doubling every 10 days', size=3) +
  annotate("text", x=0.1, y=6.5, label='Created by @eastonrwhite', size=3,hjust = 0) +
  annotate("text", x=0.1, y=6.3, label='Data from John Hopkins University/CDC', size=3,hjust = 0) 
states
  
#ggsave(filename = paste('covid_by_state_daily_updates/state_responses',substr(Sys.time(),0,10),'.jpeg',sep=''),plot = states,device = 'jpeg',width = 10,height=6,units='in',dpi=500)

#library(plotly)
#gg <- ggplotly(states, tooltip = "Province.States")
#highlight(gg, defaultValues = "Florida")
```



```{r measurements_by_states_map,echo=F,warning=F,message=F, fig.height=5,eval=T,fig.cap='Doubling time (in number of days) across the US states that recorded more than 25 cases over at least three days.\\label{fig:measurements_by_states_map}'}

# Which states have largest doubling time???

covid_by_state_doubling_time <- covid_by_state %>%
  group_by(Province.State) %>%
  filter(length(DayOfYear)>4) %>%
  summarize(exponential_param = calc_slope(Cases), doubling_time = log(2)/calc_slope(Cases), DayOf25 = mean(as.numeric(format(first_confirmed,'%j')))) %>% 
  arrange(doubling_time)

# Add back original data frame
covid_by_state_doubling_time <- left_join(covid_by_state_doubling_time, covid_by_state)

#write.csv(x = covid_by_state_doubling_time %>% group_by(Province.State) %>% filter(DayOfYear==min(DayOfYear)),
#          file = paste('covid_by_state_daily_updates/covid_responses_by_state',substr(Sys.time(),1,10),'.csv',sep=''))

# Add vector for state abbreviations
require(usmap)
state_pop <- statepop
names(state_pop)[3] <- 'Province.State'
covid_by_state_doubling_time <- left_join(covid_by_state_doubling_time,state_pop)


doubling_map <- plot_usmap(data = covid_by_state_doubling_time, values = "doubling_time", labels=FALSE) + scale_fill_viridis(begin=0.1,end=0.9) +
  theme(legend.position = "right") + 
  theme(panel.background = element_rect(color = "black")) + 
  labs(caption = "Created by: @eastonrwhite, Data from John Hopkins University/CDC",fill='Doubling time (days)',title=paste('Doubling time (days) as of',substr(Sys.time(),1,10)))
  #labs(fill='Doubling time (days)',title=paste('Doubling time (days) as of',substr(Sys.time(),1,10)))
doubling_map

#ggsave(filename = paste('covid_by_state_daily_updates/doubling_map',substr(Sys.time(),0,10),'.jpeg',sep=''),plot = doubling_map,device = 'jpeg',width = 10,height=6,units='in',dpi=500)

```


## Predictors of early (pre-intervention) state-level trajectories - NEED TO WORK ON THIS NEXT

Each US state varies considerably across a number of important axes: wealth, access to healthcare, number of international travelers, age distribution, population density, among other factors. In addition, much of the response to COVID-19 has been done at the state, as opposed to federal, government level in the US (citations). These responses have varied a lot with different laws and such... (citations). 

We had two hypotheses to explain the state-level variation in COVID-19 trajectories before government intervention occurred: human demographics as well as wealth and education indicators.



```{r, echo=F}
# Doubling time early on between states

```

```{r,early_doubling_by_predictors,echo=F,warning=F,message=F, fig.height=5,eval=F,fig.cap='Doubling time (in number of days) across the US states that recorded more than 25 cases over at least three days.\\label{fig:early_doubling_by_predictors}'}
# Combine doubling time with demographic and other indicators 

```

```{r,early_doubling_by_testing,echo=F,warning=F,message=F, fig.height=5,eval=F,fig.cap='Doubling time (in number of days) across the US states that recorded more than 25 cases over at least three days.\\label{fig:early_doubling_by_testing}'}
# Combine doubling time with information on testing

```









## Predictors of overall state-level trajectories

We had three hypotheses to explain the state-level variation in COVID-19 trajectories: human demographics, wealth and education indicators, and governmental actions. 


```{r, overall_doubling, echo=F}
# Update to date doubling time between states
covid_by_state <- covid %>%
  group_by(Date,Province.State) %>%
  summarise(Cases = sum(Cases),Deaths=sum(Deaths),Recovered = sum(Recovered), DayOfYear = mean(DayOfYear)) %>%
  filter(Cases > 24) 

# Record first date of more than 25 cases
time_of_25plus_cases <- covid_by_state %>%
  group_by(Province.State) %>%
  summarize(first_confirmed = Date[which(Cases>24)[1]])

covid_by_state = left_join(covid_by_state,time_of_25plus_cases,by="Province.State")


#starting_number = covid_by_state %>%
#  group_by(Province.State) %>%
#  summarize(starting_value = min(Cases))
#covid_by_state <- left_join(covid_by_state,starting_number)

covid_by_state$DaysSince25 <- as.numeric(as.Date(covid_by_state$Date,format='%m.%d.%y')-as.Date(covid_by_state$first_confirmed,format='%m.%d.%y'))

covid_by_state_doubling_time <- covid_by_state %>%
  group_by(Province.State) %>%
  summarize(exponential_param = calc_slope(Cases), doubling_time = log(2)/calc_slope(Cases), DayOf25 = mean(as.numeric(format(first_confirmed,'%j')))) %>% 
  arrange(doubling_time)

```



```{r,overall_doubling_by_predictors,echo=F,warning=F,message=F,eval=T,fig.cap='Doubling time (in number of days)....\\label{fig:overall_doubling_by_predictors}'}
# Combine doubling time with demographic and other indicators 

# Set up data frame of demographic information
require(usmap)
state_demo <- statepop
names(state_demo)[3] <- 'Province.State'

# Add information on population density, rural vs urban
state_demo <- left_join(state_demo,read.csv('US_states_correlates/PctUrbanRural_State.csv',header=T))

# Add information on age of population
state_demo <- left_join(state_demo,read.csv('US_states_correlates/percent65plus.csv',header=T))

# Add information on Vaccine and Volunteer rates
state_demo <- left_join(state_demo,read.csv('US_states_correlates/US_State_VaccinationRates.csv',header=T))
state_demo <- left_join(state_demo,read.csv('US_states_correlates/US_State_VolunteerRates.csv',header=T))

# Add wealth and schooling indicators
state_demo <- left_join(state_demo,read.csv('US_states_correlates/GDL-Indicators-(2018)-data.csv',header=T))

# Add timing of various state-level government actions
state_demo <- left_join(state_demo,read.csv('US_states_correlates/States_by_ActionsApr3.csv',header=T,stringsAsFactors = F))
state_demo$StateOfEmergency = as.numeric(format(as.Date(state_demo$StateOfEmergency,format = "%m/%d/%y"),format='%j'))
state_demo$LimitGatherings = as.numeric(format(as.Date(state_demo$LimitGatherings,format = "%m/%d/%y"),format='%j'))
state_demo$ClosePublicSchools = as.numeric(format(as.Date(state_demo$ClosePublicSchools,format = "%m/%d/%y"),format='%j'))
state_demo$RestrictBusinesses = as.numeric(format(as.Date(state_demo$RestrictBusinesses,format = "%m/%d/%y"),format='%j'))
state_demo$StayAtHome = as.numeric(format(as.Date(state_demo$StayAtHome,format = "%m/%d/%y"),format='%j'))

all_data_by_state <- left_join(state_demo,covid_by_state_doubling_time)

# Build plots of various demographic and indicator correlates

par(mfrow=c(4,3),mar=c(4,0.3,1,0),oma=c(0,5,0.5,0.5))

# Density factors
plot(log(all_data_by_state$POP_ST/all_data_by_state$AREA_ST),all_data_by_state$doubling_time,pch=16,las=1,ylab='',xlab='log (Population density)')
plot(all_data_by_state$POPDEN_URBAN,all_data_by_state$doubling_time,pch=16,las=1,ylab='',xlab='Population density in urban areas')
plot(all_data_by_state$POPDEN_RURAL,all_data_by_state$doubling_time,pch=16,las=1,ylab='',xlab='Population density in rural areas') #MAYBE?
plot(all_data_by_state$POPPCT_URBAN,all_data_by_state$doubling_time,pch=16,las=1,ylab='',xlab='Population percent in Urban')
plot(all_data_by_state$POPDEN_UC,all_data_by_state$doubling_time,pch=16,las=1,ylab='',xlab='Population percent in Urban')

# Health and schooling
plot(all_data_by_state$Life_expectancy,all_data_by_state$doubling_time,pch=16,las=1,ylab='',xlab='Life expectancy (years)')
plot(all_data_by_state$Percent65_plus,all_data_by_state$doubling_time,pch=16,las=1,ylab='',xlab='Percent of population above 65+')
plot(all_data_by_state$GNI_per_capita_inthousands_of_USD_2011PPP.,all_data_by_state$doubling_time,pch=16,las=1,ylab='',xlab='Gross income per capita (1000s of USD)')
plot(all_data_by_state$Expected_years_schooling,all_data_by_state$doubling_time,pch=16,las=1,ylab='',xlab='Expected years of schooling')

# Other indicators
plot(all_data_by_state$Vaccination_Rate,all_data_by_state$doubling_time,pch=16,las=1,ylab='',xlab='Yearly flu vaccination rate')
plot(all_data_by_state$VolunteerRate2015_Americorps,all_data_by_state$doubling_time,pch=16,las=1,ylab='',xlab='Volunteer rate')

mtext(text = 'Doubling time',side = 2,line = 2,outer = T,cex=1.2)

```

We found that demography, education, and wealth were poor predictors of the state-level doubling times (Fig. \ref{fig:overall_doubling_by_predictors}). Therefore, we also examined the correlation between doubling time and state government interventions, and timing of those interventions. We specifically looked at whether or not a state had implemented a specific action by the first day they had 25 or more cases. We looked at closing of public schools, limiting large gatherings (usually of more than 10 people), restricting business, and stay at home orders. Of these factors, only restricting businesses was a significant predictor of doubling time (Fig. \ref{fig:overall_doubling_by_actions}). 


```{r,overall_doubling_by_actions,echo=F,warning=F,message=F, fig.height=5,eval=T,fig.cap='Doubling time (in number of days) across the US states that recorded more than 25 cases over at least three days.\\label{fig:overall_doubling_by_actions}'}
# Combine doubling time with government actions and build plots


# Create binary data 

#all_data_by_state$StateOfEmergency <- as.numeric(all_data_by_state$StateOfEmergency<all_data_by_state$DayOf25) + 1
all_data_by_state$LimitGatherings <- as.numeric(all_data_by_state$LimitGatherings<all_data_by_state$DayOf25) 
all_data_by_state$LimitGatherings[is.na(all_data_by_state$LimitGatherings)] = 0 
all_data_by_state$ClosePublicSchools <- as.numeric(all_data_by_state$ClosePublicSchools<all_data_by_state$DayOf25) 
all_data_by_state$ClosePublicSchools[is.na(all_data_by_state$ClosePublicSchools)] = 0 
all_data_by_state$RestrictBusinesses <- as.numeric(all_data_by_state$RestrictBusinesses<all_data_by_state$DayOf25) 
all_data_by_state$RestrictBusinesses[is.na(all_data_by_state$RestrictBusinesses)] = 0
all_data_by_state$StayAtHome <- as.numeric(all_data_by_state$StayAtHome<all_data_by_state$DayOf25) 
all_data_by_state$StayAtHome[is.na(all_data_by_state$StayAtHome)] =0

all_data_by_state$total_score = (all_data_by_state$LimitGatherings) + (all_data_by_state$ClosePublicSchools) + (all_data_by_state$RestrictBusinesses) + (all_data_by_state$StayAtHome) 



# Build plot
par(mfrow=c(2,2),mar=c(4,0.3,1,0),oma=c(0,5,0.5,0.5))
#boxplot(all_data_by_state$doubling_time~all_data_by_state$StateOfEmergency,las=1)
boxplot(all_data_by_state$doubling_time~all_data_by_state$LimitGatherings,las=1,xlab='Limit gatherings by day of 25 cases',xaxt='n')
axis(side = 1,at = 1:2,labels = c('No','Yes'))
points(all_data_by_state$LimitGatherings+1,all_data_by_state$doubling_time,pch=16)
boxplot(all_data_by_state$doubling_time~all_data_by_state$ClosePublicSchools,las=1,xlab='Close schools by day of 25 cases',xaxt='n',yaxt='n')
axis(side = 1,at = 1:2,labels = c('No','Yes'))
points(all_data_by_state$ClosePublicSchools+1,all_data_by_state$doubling_time,pch=16)
boxplot(all_data_by_state$doubling_time~all_data_by_state$RestrictBusinesses,las=1,xlab='Restrict businesses by day of 25 cases',xaxt='n')
text(2,5.2,labels = '*significant difference')
axis(side = 1,at = 1:2,labels = c('No','Yes'))
points(all_data_by_state$RestrictBusinesses+1,all_data_by_state$doubling_time,pch=16)
boxplot(all_data_by_state$doubling_time~all_data_by_state$StayAtHome,las=1,xlab='Stay at home order by day of 25 cases',xaxt='n',yaxt='n')
axis(side = 1,at = 1:2,labels = c('No','Yes'))
points(all_data_by_state$StayAtHome+1,all_data_by_state$doubling_time,pch=16)

mtext(text = 'Doubling time (days)',side = 2,outer = T,line=2.5,cex.lab=1.2)


# significance tests - ask Emily

#with(all_data_by_state,summary(lm(doubling_time~RestrictBusinesses)))


```


```{r, overall_doubling_by_testkits,echo=F,warning=F,message=F, fig.height=5,eval=T,fig.cap='Doubling time (in number of days) across the US states that recorded more than 25 cases over at least three days.\\label{fig:overall_doubling_by_testkits}'}

# Test kit data over time with sensitivity
testkits <- read.csv('http://covidtracking.com/api/states/daily.csv',header=T)
testkits$date <- as.Date(as.character(testkits$date),'%Y%m%d')
testkits$percent_positive <- testkits$positive/testkits$total

require(usmap)
state_pop <- statepop[,2:4]
names(state_pop)[1] <- 'state'
state_pop = rbind(state_pop,c('VI','U.S. Virgin Islands',107710))
state_pop = rbind(state_pop,c('PR','Puerto Rico',3473000))
state_pop = rbind(state_pop,c('GU','Guam',161797))
state_pop = rbind(state_pop,c('AS','American Samoa',55537))
state_pop = rbind(state_pop,c('MP','Northern Mariana Islands',54816 ))
state_pop$pop_2015 <- as.numeric(state_pop$pop_2015)

testkits <- left_join(testkits,state_pop)

ranking_states_by_tests <- testkits %>%
  filter(date == max(date)) %>%
  rename(abbr = state) %>%
  group_by(abbr) %>%
  summarize(testkits_per_capita = total/pop_2015) %>% arrange(testkits_per_capita)


all_data_by_state2 <- left_join(all_data_by_state,ranking_states_by_tests)

plot(all_data_by_state2$testkits_per_capita,all_data_by_state2$doubling_time,las=1,pch=16,ylab='Doubling time',xlab='Total test kits per capita',cex.lab=1.2,xlim=c(0,1.1*max(all_data_by_state2$testkits_per_capita)))
text(all_data_by_state2$testkits_per_capita,all_data_by_state2$doubling_time,labels = all_data_by_state2$abbr,cex=0.7,pos=4)

```



# Conclusions and Future Work

We found a large degree of heterogeneity in the number of cases over time across US states. After state-level government actions were implemented, doubling time was most strongly correlated to restrictions on businesses. More detailed work will be needed to understand how these dynamics differ within each state, especially as many government actions started on these more local scales.








# Code availability and acknowledglements

All code and corresponding data is freely available at https://github.com/eastonwhite/COVID19_US_States. The original raw data has been compiled by the Johns Hopkins University Center for Systems Science and Engineering at (https://github.com/CSSEGISandData/COVID-19). The authors received no specific funding for this work.

# References